---
title: "Interactive Data Exploration with Plotly"
author: "Miguel Rosa"
date: "July 24, 2026"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: flatly
    highlight: tango
    self_contained: true
---

```{r setup, include=FALSE}
library(plotly)
library(dplyr)
data(mtcars)
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$am  <- ifelse(mtcars$am == 1, "Manual", "Automatic")
mtcars$vs  <- ifelse(mtcars$vs == 1, "Straight", "V-shaped")
```

---

## The `mtcars` Dataset

_32 cars. 11 variables. One goal: understand fuel economy._

```{r summary-table, echo=FALSE, results='asis'}
cat("
| Statistic | MPG | Weight (1000 lbs) | Horsepower |
|-----------|-----|-------------------|------------|
| Min       | ", min(mtcars$mpg), " | ", min(mtcars$wt), " | ", min(mtcars$hp), " |
| Median    | ", median(mtcars$mpg), " | ", median(mtcars$wt), " | ", median(mtcars$hp), " |
| Max       | ", max(mtcars$mpg), " | ", max(mtcars$wt), " | ", max(mtcars$hp), " |
")
```

---

## 3D Scatter: MPG ~ Weight + Horsepower

A 360-degree view of how weight and horsepower jointly affect fuel economy. **Drag to rotate.** Hover for details.

```{r scatter3d, echo=FALSE, warning=FALSE}
plot_ly(
  mtcars,
  x = ~wt,
  y = ~hp,
  z = ~mpg,
  color = ~cyl,
  colors = c("#E41A1C", "#4DAF4A", "#377EB8"),
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 6),
  text = ~paste(
    "<b>", rownames(mtcars), "</b><br>",
    "MPG:", round(mpg, 1), "<br>",
    "Weight:", wt, "× 1000 lbs<br>",
    "HP:", hp, "<br>",
    "Cylinders:", cyl, "<br>",
    "Transmission:", am
  ),
  hoverinfo = "text"
) %>%
  layout(
    title = "MPG by Weight & Horsepower",
    scene = list(
      xaxis = list(title = "Weight (1000 lbs)"),
      yaxis = list(title = "Horsepower"),
      zaxis = list(title = "MPG")
    )
  )
```

---

## What Drives MPG? A Look at Correlations

Highlighted bars show each variable's correlation with MPG — blue for positive, red for negative.

```{r corrplot, echo=FALSE, warning=FALSE}
num_vars <- c("mpg", "disp", "hp", "drat", "wt", "qsec")
cor_data <- mtcars %>%
  select(all_of(num_vars)) %>%
  mutate(across(everything(), as.numeric))
cor_vec <- cor(cor_data)[1, -1]
cor_df <- data.frame(
  variable = names(cor_vec),
  correlation = round(cor_vec, 3),
  sign = ifelse(cor_vec >= 0, "Positive", "Negative")
)

plot_ly(
  cor_df,
  x = ~correlation,
  y = ~reorder(variable, correlation),
  type = "bar",
  orientation = "h",
  color = ~sign,
  colors = c("#E41A1C", "#377EB8"),
  text = ~paste(correlation),
  textposition = "outside",
  hoverinfo = "text",
  marker = list(line = list(color = "rgba(0,0,0,0.3)", width = 1))
) %>%
  layout(
    title = "Correlation with MPG",
    xaxis = list(title = "Pearson Correlation", range = c(-1, 1)),
    yaxis = list(title = ""),
    showlegend = FALSE,
    margin = list(l = 80)
  )
```

---

## Transmission & MPG: Side-by-Side

```{r boxplot, echo=FALSE, warning=FALSE}
plot_ly(
  mtcars,
  x = ~am,
  y = ~mpg,
  color = ~am,
  colors = c("#E41A1C", "#377EB8"),
  type = "box",
  boxpoints = "all",
  jitter = 0.3,
  pointpos = -1.8,
  text = ~paste(rownames(mtcars), "<br>MPG:", round(mpg, 1)),
  hoverinfo = "text"
) %>%
  layout(
    title = "MPG Distribution by Transmission Type",
    xaxis = list(title = "Transmission"),
    yaxis = list(title = "MPG"),
    showlegend = FALSE
  )
```

---

## Parallel Coordinates: See the Full Picture

Each line is a car. Drag axes to filter and explore.

```{r parallel, echo=FALSE, warning=FALSE}  
df_p <- mtcars
df_p$mpg  <- as.numeric(df_p$mpg)
df_p$disp <- as.numeric(df_p$disp)
df_p$hp   <- as.numeric(df_p$hp)
df_p$wt   <- as.numeric(df_p$wt)
df_p$qsec <- as.numeric(df_p$qsec)

plot_ly(
  df_p,
  type = "parcoords",
  line = list(
    color = ~as.numeric(mtcars$cyl),
    colorscale = list(
      list(0, "#E41A1C"),
      list(0.5, "#4DAF4A"),
      list(1, "#377EB8")
    )
  ),
  dimensions = list(
    list(label = "MPG",        values = ~mpg,  range = c(10, 35)),
    list(label = "Displacement", values = ~disp, range = c(70, 475)),
    list(label = "HP",         values = ~hp,   range = c(50, 340)),
    list(label = "Weight",    values = ~wt,   range = c(1.5, 5.5)),
    list(label = "1/4 Mile",  values = ~qsec, range = c(14, 24))
  )
) %>%
  layout(title = "Parallel Coordinates — Color = Cylinders")
```

---

## Built With

| Tool | Purpose |
|------|---------|
| **R** | Data analysis & wrangling |
| **plotly** | Interactive visualizations |
| **rmarkdown** | Document generation |
| **GitHub Pages** | Hosting |

_Date: July 24, 2026_