---
title: "Next-Word Predictor"
subtitle: "A Stupid Backoff n-gram model for predictive text"
author: "Miguel Rosa"
date: "`r Sys.Date()`"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(data.table)
library(ggplot2)

results <- readRDS("../models/evaluation_results.rds")
get_metric <- function(name) results[metric == name, value]
```

## The Problem

- Mobile keyboards need to suggest the next word **instantly** as you type.
- Built on ~3.8M lines of blogs, news, and tweets (SwiftKey corpus).
- Constraints that matter as much as accuracy:
    - **Size** — must fit comfortably in a phone's / free-tier server's memory.
    - **Speed** — a suggestion that arrives after you've already typed the word is useless.

## How the Model Works

- **Stupid Backoff**: predict from the longest context we've seen, and fall back
  to shorter contexts when we haven't.

```
"I want to go to the"  →  try 3-word context "go to the"
                          not enough matches? → try "to the"
                          still not enough?   → try "the"
                          still not enough?   → most common words overall
```

- Each backoff step is discounted (×0.4) so longer, more specific contexts are
  always preferred when available.
- Rare words (seen only once) are folded into a shared `<unk>` token before
  counting — this bounds vocabulary size without losing the context they
  provided.
- Only the top 5 candidate words are kept per context — predictions never need
  more, and this is most of the memory savings.

## Performance

```{r perf-table}
knitr::kable(results, col.names = c("Metric", "Value"), digits = 2)
```

```{r perf-chart, fig.height=3.2}
acc <- results[metric %in% c("accuracy@1", "accuracy@2", "accuracy@3")]
acc[, metric := factor(metric, levels = c("accuracy@1", "accuracy@2", "accuracy@3"))]
ggplot(acc, aes(x = metric, y = value)) +
  geom_col(fill = "#2980b9", width = 0.6) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(x = NULL, y = "Accuracy", title = "Next-word accuracy on held-out test data") +
  theme_minimal(base_size = 13)
```

## The App

- A Shiny app takes a phrase and shows the top 3 predicted next words as
  clickable buttons — click one to append it and keep typing.
- Predictions update live as you type (debounced ~300ms).
- Try it: **[miguelrzazo.shinyapps.io/next-word-predictor](https://miguelrzazo.shinyapps.io/next-word-predictor/)**

## Takeaways & Next Steps

- Stupid Backoff trades a small amount of accuracy for simplicity and speed
  versus Kneser-Ney smoothing — no discount estimation needed, and lookups
  are O(1) keyed joins.
- Limitations: doesn't use semantic similarity, and rare/unseen phrasing still
  falls back all the way to generic word frequency.
- Future improvements: word embeddings for semantic backoff, user-specific
  adaptation, larger training sample if size/latency budget allows.
