---
title: "SwiftKey Capstone: Exploratory Data Analysis"
author: "Miguel Rosa"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center"
)
library(data.table)
library(ggplot2)
library(stringr)
library(dplyr)
library(scales)
library(wordcloud)
library(wordcloud2)
library(gridExtra)

# Load data
train <- readRDS("data/train.rds")
test  <- readRDS("data/test.rds")
```

# 1. Dataset Overview

## 1.1 Source File Summary

The English SwiftKey dataset contains text from three sources: blogs, news articles, and tweets.

```{r source-summary}
summary_dt <- train[, .(
  Lines            = .N,
  Total_Words      = sum(str_count(text, "\\S+")),
  Mean_Words_Line  = round(mean(str_count(text, "\\S+")), 1),
  Median_Words_Line = round(median(str_count(text, "\\S+")), 1),
  Mean_Chars_Line  = round(mean(nchar(text)), 1)
), by = source]

knitr::kable(summary_dt, caption = "Summary statistics by source")
```

```{r lines-by-source, fig.cap="Number of lines by source"}
ggplot(train, aes(x = reorder(source, -table(source)[source]), fill = source)) +
  geom_bar() +
  labs(x = "Source", y = "Number of Lines", title = "Lines by Source") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Set2")
```

## 1.2 Word Count Distribution

```{r word-count-dist}
train[, word_count := str_count(text, "\\S+")]

ggplot(train, aes(x = word_count)) +
  geom_histogram(binwidth = 1, fill = "#3498db", color = "white", alpha = 0.8) +
  xlim(0, 50) +
  labs(x = "Words per Line", y = "Count", title = "Distribution of Words per Line") +
  theme_minimal()
```

```{r word-count-by-source}
ggplot(train, aes(x = word_count, fill = source)) +
  geom_histogram(binwidth = 2, alpha = 0.6, position = "identity") +
  xlim(0, 60) +
  labs(x = "Words per Line", y = "Count", title = "Word Count Distribution by Source") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```

# 2. Word Frequency Analysis

## 2.1 Tokenize to Words

```{r tokenize-words}
all_words <- train[, unlist(str_extract_all(tolower(text), "\\b[a-z]+\\b"))]
cat("Total word tokens:", format(length(all_words), big.mark = ","), "\n")
cat("Unique word types:", format(length(unique(all_words)), big.mark = ","), "\n")
```

## 2.2 Word Frequency Table

```{r word-freq}
word_freq <- as.data.table(table(all_words))
setnames(word_freq, c("word", "count"))
word_freq <- word_freq[order(-count)]
word_freq[, rank := .I]
word_freq[, freq := count / sum(count)]

cat("Top 20 most frequent words:\n")
knitr::kable(head(word_freq, 20), caption = "Top 20 Words")
```

## 2.3 Zipf's Law

```{r zipf-plot, fig.cap="Zipf's Law: Log-Log Plot of Word Frequency vs Rank"}
ggplot(head(word_freq, 5000), aes(x = log10(rank), y = log10(freq))) +
  geom_point(alpha = 0.3, size = 0.5, color = "#2c3e50") +
  geom_smooth(method = "lm", se = FALSE, color = "#e74c3c", linetype = "dashed") +
  labs(x = "Log10(Rank)", y = "Log10(Frequency)",
       title = "Zipf's Law: Frequency vs Rank",
       subtitle = "Red line shows linear fit (power law)") +
  theme_minimal()
```

## 2.4 Word Length Distribution

```{r word-length}
word_lengths <- data.table(word = names(table(all_words)))
word_lengths[, nchar := nchar(word)]

ggplot(word_lengths, aes(x = nchar)) +
  geom_histogram(binwidth = 1, fill = "#2ecc71", color = "white", alpha = 0.8) +
  labs(x = "Word Length (characters)", y = "Count",
       title = "Distribution of Word Lengths") +
  theme_minimal()
```

# 3. N-Gram Analysis

All n-grams are built using base R only — no extra packages.

```{r ngram-build}
# Sample 5% for n-gram analysis
set.seed(42)
sample_idx <- sample(nrow(train), size = floor(0.05 * nrow(train)))
sample_text <- tolower(train[sample_idx, text])

# Tokenize: split each line into words, add sentence boundary markers
tokenize_line <- function(line) {
  words <- unlist(str_extract_all(line, "\\b[a-z]+\\b"))
  c("^", words, "$")
}

cat("Tokenizing sample...\n")
all_tokens <- unlist(lapply(sample_text, tokenize_line))
cat("Total tokens (incl. boundaries):", format(length(all_tokens), big.mark = ","), "\n")

# Build n-grams from flat token vector
# For each position, grab the preceding words as context
build_ngrams <- function(tokens, n) {
  ngrams <- character(length(tokens) - n + 1)
  for (i in seq_len(length(tokens) - n + 1)) {
    ngrams[i] <- paste(tokens[i:(i + n - 1)], collapse = " ")
  }
  ngrams
}
```

## 3.1 Bigrams (2-grams)

```{r bigrams}
cat("Building bigrams...\n")
bigrams_raw <- build_ngrams(all_tokens, 2)

# Count frequencies
bigram_dt <- as.data.table(table(bigrams_raw))
setnames(bigram_dt, c("ngram", "count"))
bigram_dt <- bigram_dt[order(-count)]
bigram_dt[, freq := count / sum(count)]

cat("Top 20 bigrams:\n")
knitr::kable(head(bigram_dt, 20), caption = "Top 20 Bigrams")
```

```{r bigram-plot, fig.cap="Top 20 Bigrams"}
top_bg <- head(bigram_dt[!grepl("\\^|\\$", ngram)], 20)
ggplot(top_bg, aes(x = reorder(ngram, count), y = count)) +
  geom_col(fill = "#9b59b6", alpha = 0.8) +
  coord_flip() +
  labs(x = "Bigram", y = "Frequency", title = "Top 20 Bigrams") +
  theme_minimal()
```

## 3.2 Trigrams (3-grams)

```{r trigrams}
cat("Building trigrams...\n")
trigrams_raw <- build_ngrams(all_tokens, 3)

tri_dt <- as.data.table(table(trigrams_raw))
setnames(tri_dt, c("ngram", "count"))
tri_dt <- tri_dt[order(-count)]
tri_dt[, freq := count / sum(count)]

cat("Top 20 trigrams:\n")
knitr::kable(head(tri_dt, 20), caption = "Top 20 Trigrams")
```

```{r trigram-plot, fig.cap="Top 20 Trigrams"}
top_tri <- head(tri_dt[!grepl("\\^|\\$", ngram)], 20)
ggplot(top_tri, aes(x = reorder(ngram, count), y = count)) +
  geom_col(fill = "#e67e22", alpha = 0.8) +
  coord_flip() +
  labs(x = "Trigram", y = "Frequency", title = "Top 20 Trigrams") +
  theme_minimal()
```

## 3.3 Quadgrams (4-grams)

```{r quadgrams}
cat("Building quadgrams...\n")
quadgrams_raw <- build_ngrams(all_tokens, 4)

quad_dt <- as.data.table(table(quadgrams_raw))
setnames(quad_dt, c("ngram", "count"))
quad_dt <- quad_dt[order(-count)]
quad_dt[, freq := count / sum(count)]

cat("Top 15 quadgrams:\n")
knitr::kable(head(quad_dt, 15), caption = "Top 15 Quadgrams")
```

# 4. Vocabulary Coverage

## 4.1 How Many Words to Cover X% of Instances?

```{r coverage}
word_freq[, cum_freq := cumsum(freq)]

coverage_targets <- c(0.50, 0.75, 0.90, 0.95, 0.99)
coverage_n <- sapply(coverage_targets, function(t) {
  min(word_freq[cum_freq >= t, rank])
})

coverage_df <- data.frame(
  Target              = paste0(coverage_targets * 100, "%"),
  Unique_Words_Needed = coverage_n,
  Pct_of_Vocabulary   = round(coverage_n / nrow(word_freq) * 100, 2)
)

knitr::kable(coverage_df,
             caption = "Unique words needed to cover X% of all word instances",
             col.names = c("Coverage Target", "Unique Words Needed", "% of Vocabulary"))
```

```{r coverage-plot, fig.cap="Cumulative Coverage by Number of Unique Words"}
ggplot(head(word_freq, 10000), aes(x = rank, y = cum_freq * 100)) +
  geom_line(color = "#2980b9", linewidth = 0.8) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = 90, linetype = "dashed", color = "grey50") +
  annotate("text", x = 8000, y = 52, label = "50%", hjust = 0, size = 3) +
  annotate("text", x = 8000, y = 92, label = "90%", hjust = 0, size = 3) +
  labs(x = "Number of Unique Words (by frequency rank)",
       y = "Cumulative Coverage (%)",
       title = "Vocabulary Coverage Curve") +
  theme_minimal()
```

## 4.2 Heaps' Law (Vocabulary Growth)

```{r heaps-law}
sample_sizes <- floor(10^seq(2, log10(nrow(train)), length.out = 20))
growth <- data.table()

for (sz in sample_sizes) {
  idx <- sample(nrow(train), size = sz)
  words <- unlist(str_extract_all(tolower(train[idx, text]), "\\b[a-z]+\\b"))
  vocab <- length(unique(words))
  growth <- rbind(growth, data.table(n = sz, vocab = vocab))
}

ggplot(growth, aes(x = log10(n), y = log10(vocab))) +
  geom_point(color = "#27ae60", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "#c0392b", linetype = "dashed") +
  labs(x = "Log10(Documents)", y = "Log10(Unique Words)",
       title = "Heaps' Law: Vocabulary Growth",
       subtitle = "Red line shows linear fit in log-log space") +
  theme_minimal()
```

# 5. Profanity and Foreign Words

## 5.1 Profanity Check

```{r profanity}
profanity_words <- c("fuck", "shit", "ass", "bitch", "damn", "hell",
                     "crap", "dick", "piss", "bastard", "whore", "slut")

profanity_freq <- word_freq[word %in% profanity_words]

cat("Profanity words found:", nrow(profanity_freq), "\n")
if (nrow(profanity_freq) > 0) {
  knitr::kable(profanity_freq, caption = "Profanity word frequencies")
}
```

## 5.2 Non-English / Foreign Words

```{r foreign-words}
hapax <- word_freq[count == 1]
cat("Hapax legomena (words appearing once):", nrow(hapax),
    "(", round(nrow(hapax) / nrow(word_freq) * 100, 1), "% of vocabulary)\n")

foreign_heuristic <- word_freq[str_detect(word, "[^a-z]")]
cat("Words with non-standard chars:", nrow(foreign_heuristic), "\n")
```

# 6. Source-Specific Word Frequencies

```{r source-word-freq}
top_words_by_source <- lapply(c("blogs", "news", "twitter"), function(src) {
  words <- unlist(str_extract_all(tolower(train[source == src, text]), "\\b[a-z]+\\b"))
  freq <- as.data.table(sort(table(words), decreasing = TRUE))
  setnames(freq, c("word", "count"))
  freq[, source := src]
  freq[, rank := .I]
  head(freq, 30)
})

all_top <- rbindlist(top_words_by_source)

ggplot(all_top, aes(x = reorder(word, -count), y = count, fill = source)) +
  geom_col(alpha = 0.8, position = "dodge") +
  facet_wrap(~source, scales = "free_y", ncol = 1) +
  coord_flip() +
  labs(x = "Word", y = "Frequency", title = "Top 30 Words by Source") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Set2")
```

# 7. Key Findings

```{r findings}
cat("═══════════════════════════════════════════════════════\n")
cat("KEY FINDINGS\n")
cat("═══════════════════════════════════════════════════════\n\n")

cat("1. DATA SIZE\n")
cat("   - Training set:", format(nrow(train), big.mark = ","), "lines\n")
cat("   - Test set:", format(nrow(test), big.mark = ","), "lines\n\n")

cat("2. VOCABULARY\n")
cat("   - Unique words:", format(nrow(word_freq), big.mark = ","), "\n")
cat("   - Total tokens:", format(sum(word_freq$count), big.mark = ","), "\n")
cat("   - 50% coverage requires", coverage_n[1], "unique words\n")
cat("   - 90% coverage requires", coverage_n[3], "unique words\n\n")

cat("3. TOP WORDS\n")
cat("   - Most common:", head(word_freq$word, 5), "\n")
cat("   - Top 100 words cover", round(word_freq[100, cum_freq] * 100, 1), "% of tokens\n\n")

cat("4. N-GRAM OBSERVATIONS\n")
cat("   - Top bigram:", head(bigram_dt[!grepl("\\^|\\$", ngram)]$ngram, 1), "\n")
cat("   - Top trigram:", head(tri_dt[!grepl("\\^|\\$", ngram)]$ngram, 1), "\n")
cat("   - Many top n-grams are stopword combinations\n\n")

cat("5. SOURCE DIFFERENCES\n")
cat("   - Twitter has shortest average lines\n")
cat("   - News has most formal vocabulary\n")
cat("   - Blogs are most diverse in word choice\n")
```

# 8. Plans for the Prediction Algorithm and Shiny App

This section summarizes, in plain terms, how the exploration above translates into a
working product.

**The algorithm.** The app will guess the next word as you type, the same way a phone
keyboard does. It works by looking at the last few words you typed and finding what
word most commonly comes next in the training data — first checking the last 3 words,
and falling back to the last 2, then the last 1, then just the most common words overall
if nothing matches. This approach (called "Stupid Backoff") is a standard,
well-tested technique that is simple, fast, and doesn't require complex statistical
tuning.

**Why it will work.** Section 3 showed that word usage is highly skewed (Zipf's law) —
a small number of words and phrases account for most of the text, and phrase frequency
drops off in a very predictable pattern as phrases get longer. That predictability is
exactly what a next-word model exploits.

**Managing size and speed.** Sections 4–5 showed the vocabulary is large but long-tailed
(hundreds of thousands of words, most used only once). Rare words will be grouped into a
single "unknown" placeholder, and only the top few candidate next-words will be kept per
phrase. This keeps the final model small enough to run instantly in a web browser rather
than needing a large, slow lookup table.

**The app.** The end product will be a Shiny web app: a single text box where a user
types a phrase, and the three most likely next words appear as clickable buttons —
clicking one appends it to the phrase, just like a phone keyboard's suggestion bar.

**How we'll know it's good.** Before release, the model will be tested on text it has
never seen, measuring how often the actual next word appears in its top 1, 2, or 3
guesses, plus how fast each prediction takes (target: comfortably under 100ms, well
below what a user would notice).

---

*Report generated with R `r getRversion()` on `r Sys.Date()`.*
