---
title: "Car MPG Predictor"
subtitle: "An Interactive Shiny Application"
author: "Miguel Rosa"
date: "2026"
output:
  ioslides_presentation:
    widescreen: true
    small_font: true
---

## Why Predict MPG?

- Fuel economy affects **cost**, **environment**, and **policy**
- Estimating MPG from car specs requires statistical knowledge
- This app makes it **simple** for anyone

```{r, echo=TRUE, results='asis'}
data(mtcars)
cat("Dataset:", nrow(mtcars), "cars,", ncol(mtcars), "variables")
```

---

## The Model

Multiple linear regression using three key predictors:

$$\hat{MPG} = \beta_0 + \beta_1 \cdot Weight + \beta_2 \cdot HP + \beta_3 \cdot Cylinders$$

```{r, echo=TRUE}
data(mtcars)
fit <- lm(mpg ~ wt + hp + cyl, data = mtcars)
round(summary(fit)$coefficients, 4)
```

<br>

R-squared = **0.84** — the model explains 84% of MPG variance.

---

## How It Works

**Three simple inputs:**

1. **Weight** — 1,500 to 5,500 lbs
2. **Horsepower** — 50 to 350 HP
3. **Cylinders** — 4, 6, or 8

**One click** → instant prediction + scatterplot

```{r, echo=TRUE, fig.height=3.5, fig.width=6}
library(ggplot2)
mtcars$cyl_f <- as.factor(mtcars$cyl)
ggplot(mtcars, aes(wt, mpg, color=cyl_f)) + 
  geom_point(size=3) + 
  labs(title="Weight vs MPG by Cylinders",
       x="Weight (1000 lbs)", y="MPG", color="Cylinders") +
  theme_minimal()
```

---

## Live Demo

- Built with **R + Shiny + ggplot2**
- Model: `lm(mpg ~ wt + hp + cyl)` on the `mtcars` dataset

**Example predictions:**

```{r, echo=TRUE}
fit <- lm(mpg ~ wt + hp + cyl, data = mtcars)
examples <- data.frame(
  wt = c(2.5, 3.5, 5.0),
  hp = c(100, 150, 250),
  cyl = c(4, 6, 8)
)
preds <- round(predict(fit, examples), 1)
cat("Light / 4-cyl:", preds[1], "MPG\n")
cat("Mid   / 6-cyl:", preds[2], "MPG\n")
cat("Heavy / 8-cyl:", preds[3], "MPG\n")
```

---

## Summary

| Feature | Detail |
|---------|--------|
| **Inputs** | 3 sliders (weight, HP, cylinders) |
| **Output** | Predicted MPG + scatterplot |
| **Model** | Linear regression, R² = 0.84 |
| **Data** | `mtcars` (32 observations) |
| **Tools** | R, Shiny, ggplot2 |

Source code: [github.com/miguelrzazo/Data-Science](https://github.com/miguelrzazo/Data-Science)

**Try it out!**
